---
title: "PA1_template"
author: "KB"
date: "Friday, November 13, 2015"
output: html_document
---
#Data Science : Reproducible Research - Assignment 1

##Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a [Fitbit](http://www.fitbit.com/), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuel), or [Jawbone Up](https://jawbone.com/up). These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

##Data
The data for this assignment can be downloaded from the course web site:

- Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FActivity.zip) (52K) 

The variables included in this dataset are:

- **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- **date**: The date on which the measurement was taken in YYYY-MM-DD format
- **interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

##Assignment

The assignment consists of 5 sections

1. Loading and preprocessing the data
2. What is mean total number of steps taken per day?
3. What is the average daily activity pattern?
4. Imputing missing values
5. Are there differences in activity patterns between weekdays and weekends?

###Section 1 : Loading and preprocessing the data

*NOTE:* You must set the Working Directory to point to the folder where you saved the zip file and where you will be running this R Markdown file.

####**1. Load the data**

The link provided a zip file. Proceeding to unzip the file and then load activity.csv file as a dataframe in R.

```{r loaddata, echo=TRUE}
library(ggplot2)
unzip(zipfile="repdata-data-activity.zip")
fitdata<-read.csv("./activity.csv")
```

####**2. Process/transform the data (if necessary) into a format suitable for your analysis**

We need to capture the dates (unique dates) from the dataset and transform the date into the POSIXlt class so we can capture the julian date, so we can use it later on.

```{r transformdata, echo=TRUE}
alldates <- strptime(fitdata$date, "%Y-%m-%d")
fitdata$date <- alldates
unqdates <- unique(alldates)
unqintervals <- unique(fitdata$interval)
```

###Section 2 : What is mean total number of steps taken per day?

####**1. Calculate the total number of steps taken per day**

Calculating the number of steps grouped by julian day

```{r numsteps, echo=TRUE}
TotStepsPerDay <- split(fitdata$steps, alldates$yday)
SumTotalTotStepsPerDay <- sapply(TotStepsPerDay, sum, na.rm=TRUE)
```

####**2. Make a histogram of the total number of steps taken each day**

Plotting a histogram to show the total number of steps taken each dat from October to November in 2012.
Coloring the histogram with 7 colors to differenciate each week.

```{r plotstepsday, echo=TRUE}
#Plotting a histogram
plot(unqdates, SumTotalTotStepsPerDay, main="Histogram of total number of steps taken each day", 
     xlab="Dates in October and November,2012)", ylab="Steps", type="h", lwd=5, 
     col=c("yellow","orange","pink","red","blue","green","black"))
dev.copy(png, file="plot1.png", height=480, width=480)
```

**OBSERVATION**, activity jumps up in the last week of November, coincidentaly during the Thanksgiving week.

####**3. Calculate and report the mean and median of the total number of steps taken per day**

Calculating the mean and median of total steps per day and assigning the values to a vector.

```{r calmeanmedian, echo=TRUE}
#Calculating the meean per julian day
MeanStepsPerDay <- sapply(TotStepsPerDay, mean, na.rm=TRUE)
MeanStepsPerDay
#Calculating the median per julian day
MedianStepsPerDay <- sapply(TotStepsPerDay, median, na.rm=TRUE)
MedianStepsPerDay
```

We can observe the average median value for per day is zero.

###Section 3 : What is the average daily activity pattern?

####**1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r timeseriesplot, echo=TRUE}
#Group the number of steps by interval
AvgStepsPerInterval <- split(fitdata$steps, fitdata$interval)
#Calculate the mean of number of steps per interval
AvgStepsPerInterval <- sapply(AvgStepsPerInterval, mean, na.rm=TRUE)
#Plotting the data as a Time Series per interval
plot(unqintervals, AvgStepsPerInterval, type="l", main = "Time Series Plot", xlab="5-minute Intervals", 
     ylab="Avg Steps",col="red", lwd=2)
dev.copy(png, file="plot2.png", height=480, width=480)
```
     
####**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r findinterval, echo=TRUE}
#Getting the maximum number of steps per interval
maxStepsPerInterval <- max(AvgStepsPerInterval, na.rm=TRUE)
AvgStepsPerInterval[which.max(AvgStepsPerInterval)]
print(paste0("Interval ", as.numeric(names(which.max(AvgStepsPerInterval)))))
```

###Section 4 : Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA).
The presence of missing days may introduce bias into some calculations or summaries of the data.

####**1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

We will first calculate the number of missing vales in the dataset.

```{r missvalues, echo=TRUE}
#Variable missingvalues will contain the number of missing values
missingvalues<-sum(is.na(fitdata$steps))
#Printing value of missing values
print(paste0("Total Number of missing values ", missingvalues))
```

####**2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

The strategy we will use to fill in the missing values in the dataset is to replace all NAN values with the median valued as zero as we previously dispayed and replace all NA values with the mean of that particular 5-minute interval where the observation falls on.

```{r fillmissingval, echo=TRUE}
#Calculating the number of unique intervals
length(unqintervals)
print(paste0("Number of intervals per day: ", as.numeric(length(unqintervals))))
#Calculating the total number of days
length(unqdates)
print(paste0("Number of days: ", as.numeric(length(unqdates))))
#There are 288 intervals per day, during 61 days
#Assigning to all intervals with NAN as a median zero
MeanStepsPerDay[is.nan(MeanStepsPerDay)] <- 0
#Create new vector of 288 length to create 288 new values containing means
newmeanvector<- rep(MeanStepsPerDay, 288)
newstepsvector<-fitdata$steps
# We need to find where the NA values are
NAsteps <- is.na(newstepsvector)
#Replacing the values that are NA with the corresponding mean value
newstepsvector[NAsteps]<-newmeanvector[NAsteps]
```

####**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**

```{r newds, echo=TRUE}
#Creating a new dataset that will store the values from the new steps dataset without NA or NAN values
newfitdata<-fitdata
newfitdata$steps<-newstepsvector
```

####**4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median  total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?**

Repeating Section 2 using the new dataset, following same steps to create the histogram

```{r makeplot2, echo=TRUE}
#Splitting the data grouping steps by date
TotStepsPerDayDF2 <- split(newfitdata$steps, alldates$yday)
#Calculates the sum of total steps per day in the new data frame
SumTotalTotStepsPerDayDF2 <- sapply(TotStepsPerDayDF2, sum, na.rm=TRUE)
#Setting the plot to be combined in 2 rows and 1 column
par(mfcol=c(2,1))
# Plotting the original histogram in the first column
plot(unqdates, SumTotalTotStepsPerDay, main="Total steps taken each day (data with missing values)", 
     xlab="Dates in October and November,2012)", ylab="Steps", type="h", lwd=5, 
     col=c("yellow","orange","pink","red","blue","green","black"))
dev.copy(png, file="plot3.png", height=480, width=480)
# Plotting the new histogram in the second column
plot(unqdates, SumTotalTotStepsPerDayDF2, main="Total steps taken each day (data without missing values)", 
     xlab="Dates in October and November,2012)", ylab="Steps", type="h", lwd=5, 
     col=c("darkgoldenrod2","chocolate1","darkmagenta","firebrick4","dodgerblue4","darkgreen","black"))
dev.copy(png, file="plot4.png", height=480, width=480)
```

Do these values differ from the estimates from the first part of the assignment?

```{r analizeplot2, echo=TRUE}
#Calculating the average number of steps per day using new dataset
MeanStepsPerDayDF2 <- sapply(TotStepsPerDayDF2, mean, na.rm=TRUE)
#Printing MEAN WITH missing values
MeanStepsPerDay
#Printing MEAN WITHOUT missing values
MeanStepsPerDayDF2
#Calculating the median number of steps per day using new dataset
MedianStepsPerDayDF2 <- sapply(TotStepsPerDayDF2, median, na.rm=TRUE)
#Printing MEDIAN WITH missing values
MedianStepsPerDay
#Printing MEDIAN WITHOUT missing values
MedianStepsPerDayDF2
```

**What is the impact of imputing missing data on the estimates of the total daily number of steps?**
Looking at the charts, the only values that seem have changed are those days where the observations were missing (NA and NAN or zeros). The rest of the observations look the same.
Adding the mean into those missing values, will not change significantlly specially when calculating the mean.

###Section 5 : Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
We will need to create 2 datasets, one for weekdays and one for weekends and split the data from the original dataset.

####**1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**


```{r weekday, echo=TRUE}
# wday is an integer ranging from 0 to 6 that represents the day of the week
# 0 is for Sunday, 1 is for Monday, ..  6 for Saturday
allwdays <- alldates$wday
# Createting a numeric vector containing data 1 for a weekday and 2 for a weekend
tablewday <- rep(0, 17568) # in total there are 17568 observations
# Assigning 1 to a weekday
tablewday[allwdays >= 1 & allwdays <= 5] <- 1
# Assigning 2 to a weekend day
tablewday[allwdays == 6 | allwdays == 0] <- 2
# Creating new dataset 
tabledays <- factor(tablewday, levels=c(1,2), labels=c("Weekdays", "Weekends"))
# Create a new column that contains this factor for each day
newfitdata$weekDay <- tabledays
# Splitting up into a week days dataframe and weekend dataframe
tableWeekdays <- newfitdata[newfitdata$weekDay == "Weekdays", ]
tableWeekends <- newfitdata[newfitdata$weekDay == "Weekends", ]
#Grouping number of steps for each week day group by interval
StepsOnWeekdays <- split(tableWeekdays$steps, tableWeekdays$interval)
StepsOnWeekends <- split(tableWeekends$steps, tableWeekends$interval)
# Calculating the average number of steps per interval
meanStepsOnWeekdays <- sapply(StepsOnWeekdays, mean)
meanStepsOnWeekends <- sapply(StepsOnWeekends, mean)
```

####**2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**

```{r plotweekday, echo=TRUE}
#Plotting the data
par(mfcol=c(2,1))
plot(unqintervals, meanStepsOnWeekdays, type="l",
     main="Average number of steps per interval across all weekdays", 
     xlab="Interval", ylab="Average steps for weekdays", 
     lwd=2, col="red")
dev.copy(png, file="plot5.png", height=480, width=480)
plot(unqintervals, meanStepsOnWeekends, type="l",
     main="Average number of steps per interval across all weekends", 
     xlab="Interval", ylab="Average steps for weekends", 
     lwd=2, col="blue")
dev.copy(png, file="plot6.png", height=480, width=480)
```

**OBSERVATION :** We can observe the average number of steps increses during the weekends and decreases during weekdays. One cause would be the lack of time for a person to do an activity outside work.


